On the Minimum Core Mass for Giant Planet Formation at Wide Separations

### 2013-10-28 tex/draft/draft4.tex
edits to section 2, mostly

The core accretion hypothesis proposes that giant planets form by the accretion of gas onto a solid protoplanetary core. The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$. This value, however, depends on several factors, such as the location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids.  Motivated by ongoing direct imaging searches for giant planets, this study focuses on core mass requirements in the outer disk.
Our model investigates the growth of atmospheres around solid cores of fixed mass.  Without planetesimal accretion to heat the atmosphere, our limiting case gives the fastest possible atmospheric growth by Kelvin-Helmholtz contraction. We develop a two-layer, spherical atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk.  For different locations in the disk, we determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.   The minimum core mass required to nucleate atmosphere collapse is $\sim$$8.5 M_{\oplus}$ at 5 AU and steadily decreases to $\sim$$3.5 M_{\oplus}$ at 100 AU, for a Solar composition atmosphere with a standard dust opacity.   This trend is mainly explained by the cooler temperatures in the outer disk. Lower pressures and densities in the outer disk have a negligible effect on core mass requirements.  At all distances, decreasing the dust opacity and/or increasing the mean molecular weight of the atmosphere reduces the critical core mass.   The ongoing accretion of solids would delay or prevent atmospheric runaway growth.  Thus our low values for minimum core masses do not contradict the existence of ice giants, but emphasize the importance of the accretion history of solids.  We also develop a non-self-gravitating, analytic cooling model to help explain our numerical trends.  Additionally, this comparison reveals that self-gravity significantly affects atmospheric evolution even when the atmosphere is only $\sim$$20\%$ as massive as the core.

\end{abstract}

%Current theories of giant planet formation postulate that these planets form either through core accretion (refs), in which solid planetesimals collide and grow into a massive solid core, which then accretes a gaseous envelope, or due to a gravitational instability in the protoplanetary disk that leads to fragmentation of the disk into self-gravitating clumps (refs, inc. \citealt{dangelo11}).  %quote murray-clay, kratter 10, rafikov 05, etc.
%
%Standard core accretion models (refs) assume that the core and the atmosphere grow at the same time, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse, while the heat generated by the gravitational (Kelvin-Helmholtz) contraction of the atmosphere is neglected. These studies consider that the planet atmosphere is in steady state, in which all the luminosity due to planetesimal accretion is radiated away by the envelope, and  find that there exists a minimum (``critical'') core mass past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs. 
%
%Forming giant planets at wide separations in the disk poses theoretical challenges. On the one hand, gravitational instability generates objects that are too massive to explain the current observed properties of exoplanets (refs, inc. \citealt{rafikov05}). On the other hand, planetesimal accretion is slow at large distances in the disk, and therefore large cores may not be able to form before the dissipation of the disk (refs). I would therefore be easier if giant planets could form from smaller cores which would need less time to grow. 
%
%In this study, we show that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs. In this scenario, the planetesimal accretion rate is significantly slowed down during the gas contraction phase of the atmosphere. This reduction can arise due to dynamical clearing, or due to the core having formed in the inner parts of the disk and migrated outwards, etc. In this situation, the atmosphere evolution is dominated by the Kelvin-Helmholtz contraction of the envelope. The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation. 
%
%In our model we therefore assume that the luminosity evolution of the atmosphere is dominated by gas contraction, while the planetesimal accretion rate is negligible. As a result, the protoplanetary core has a fixed mass. We consider that the atmosphere evolves in time through stages of quasi-static equilibrium. Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and unstable hydrodynamic collapse commences. The time required for the atmosphere to grow to this stage is the characteristic growth time of the atmosphere. For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the ``critical core mass''. 
%
%We develop a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time. We aim to find the critical core mass for a giant planet to form before the dissipation of the disk.

%Kelvin-Helmholz (KH) contraction considered previously (Ikoma, PapNel05).  Say why useful.  This paper develops a simplified model of KH contraction to both elucidate the relevant physical processes and explore a wide range of disk parameter space...

%This paper is organized as follows. In Section \ref{sec:model} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere. In Section \ref{sec:coolingan}, we present a simplified analytic model that predicts the qualitative behavior of the numerical model. Results for atmospheric structure and evolution are presented in Section \ref{sec:KH}, and implications for the critical core mass are presented in Section \ref{sec:critical}.  The discussion in Section \ref{sec:neglected} addresses various approximations and neglected effects.  We summarize our findings in Section \ref{sec:conclusions}.  The appendices contain...

@@

Models of giant planet formation fall in two categories:  core accretion or gravitational instability \citep{dangelo11, youdin13}. In core accretion models, a solid core grows until it becomes massive enough to rapidly accrete gas \citep{PerCam74, mizuno78}.  Gravitational instability (GI) theories investigate the fragmentation of the protoplanetary disk into bound clumps \citep{cameron78, boss97}.  %quote murray-clay, kratter 10, rafikov 05, etc.

%Detailed core accretion studies (\citealt{boden86}, \citealt{pollack96}, \citealt{alibert05}) have shown that this giant planet formation model can be separated in three phases. In phase 1 \citep{alibert05} a solid core forms by accreting planetesimals from its feeding zone at a high rate.

@@

This work develops quasi-static models in which the core has a fixed mass and is no longer accreting solids.  By isolating the process of KH contraction, we determine the minimum timescale for runaway atmospheric growth, and thus for giant planet formation by core accretion.  Previous studies have considered this type of growth \citep{ikoma00, pn05}, which is an end-member case of phase 2 and 3 growth in more detailed quasi-static models.  Our study develops simplified cooling models for the purposes of broadly exploring parameter space -- especially in the outer disk -- and developing a more detailed understanding of atmospheric accretion.

%Full time-dependent simulations of giant planet evolution take into account the Kelvin-Helmholtz (KH) contraction of the gaseous envelope. As a result, the atmosphere is no longer in a steady state, but rather it accretes gas as it cools radiatively. In this scenario there are two main sources that contribute to the luminosity evolution of the envelope --- planetesimal accretion and gas contraction. These evolutionary models for core and atmosphere growth include time-dependent planetesimal accretion rates that are determined by the distribution of solids in the disk, as well as the planet mass and capture cross-section. This is is contrast with the static models in which the planetesimal accretion rate is usually constant with time. \citet{pollack96} find that the planet growth can be separated into three relatively well defined phases. Phase 1 is dominated by the growth of the core due to runaway planetesimal accretion, and ends when the planet's feeding zone has been depleted of solids. At this stage, planetesimal accretion can no longer balance radiative losses, and the atmosphere cools and contracts. This is phase 2, during which the envelope mass steadily increases, while core growth is no longer significant. Once the atmosphere mass becomes comparable to the core mass, growth is continually accelerated, leading to runaway gas accretion, i.e. phase 3. Phase 2 is found to be much longer than the other two phases (see also \citealt{alibert05}), hence the evolutionary timescale of the atmosphere is set by KH contraction.

%Since KH contraction dominates the envelope accretion timescale in many core accretion models, it is useful to study atmospheric growth onto a core of fixed mass (\citealt{ikoma00}, \citealt{pn05}). This regime is the one considered in this work. We explore the atmosphere evolution around fully grown cores of different masses, and determine the minimum core mass to form a giant planet before disk dissipation for a variety of disk conditions and properties of the nebular gas. We note that our models are a special case of the time-dependent studies, with the planetesimal accretion rate set to zero. Our neglect of planetesimal accretion allows us to obtain absolute lower limits on the core mass needed to form a gas giant, as additional heat sources limit the ability of the atmosphere to cool and undergo KH contraction.

@@

Our model can accurately model the growth of the atmosphere up to the crossover mass, when the atmosphere mass equals the mass of the core.   Beyond the crossover mass,  our approximate treatment of the radiative zone (explained below) breaks down.  
Since subsequent growth is a rapid runaway process, our model can investigate to good accuracy the timescale requirement of core accretion.  Our simplified treatment is also inappropriate for hot, short-period planets, where dust sublimation gives deeper radiative zones that require more detailed models.

Our main assumptions are summarized as follows:
\begin{enumerate}
\item The atmosphere is spherically symmetric and remains in hydrostatic balance during its thermal evolution.
\item The core mass and radius are fixed in evolutionary calculations, neglecting ongoing planetesimal or dust accretion.
\item At the planet's Hill radius, the atmospheric temperature and pressure match the conditions of the disk midplane.
\item The only source of planetary luminosity is the gravitational contraction of the atmosphere.  
\item Cooling of the atmosphere is dominated by the convective interior.  Thus the luminosity in the radiative exterior is held spatially constant. %In the radiative zone, luminosity generation is neglected, i.e. the luminosity is held constant.
\item A global cooling model connects independent static solutions into a time-dependent sequence.
\item A polytropic equation of state (EOS) is assumed for simplicity. %AP: define EOS here if not already
\item Dust grains provide the opacity in the radiative zone, which remains cool enough to avoid dust sublimation.
\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass, i.e. the crossover time, is a good approximation of the total time to form a gas giant.
\end{enumerate}
The remainder of this section develops our model in more detail.

@@

The characteristic length scales for protoplanetary atmospheres are crucial for choosing boundary conditions and for understanding the validity of  spherical symmetry in a disk of scale height $H\di$.  The radius of the solid core is

\begin{equation}
\label{eq:rc}
R\co \equiv \left(\frac{3 M\co}{4 \pi \rho\co}\right)^{1/3} \approx 10^{-4} \mcn{10}^{1/3} ~\text{AU},
\end{equation}
where the core mass, $M\co$, is normalized to 10 Earth masses as $\mcn{10} \equiv M\co/(10~M_\oplus)$. The core density is held fixed at $\rho\co=3.2$ g cm$^{-3}$.  We thus neglect  the detailed equation of state of the solid core \citep{fortney07}.

A planet can bind a dense atmosphere if its escape velocity exceeds the sound speed.  This criterion is satisfied inside the Bondi radius
\begin{equation}
\label{eq:RB}
\RB \equiv \frac{G M\pla}{c\di^2} \approx 0.17 \, {\mpn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU,
\end{equation}
where the enclosed planet mass, $M\pla = M\co + M_\mathrm{atm}$, includes the core and any atmosphere within the Bondi radius.  The scalings thus use $\mpn{10} \equiv M\pla /(10~M_\oplus)$.

Stellar tides dominate the planet's gravity beyond the Hill radius
\begin{equation}
\label{eq:RHill}
R_{\rm H} = \left(M\pla \over 3 M_\ast \right)^{1/3}a \approx 0.22 \, {\mpn{10}^{1/3} \, \aun{10} }~\AU ,
\end{equation}
where hydrostatic balance breaks down.  In \Eq{eq:RHill}, $M\pla$ includes the mass enclosed within $\RH$.  During the early stages of evolution, $M\pla \sim M\co$, and the core mass can be used to get estimates of both $\RB$ and $\RH$.

@@

In the low mass regime, $M\pla < M_{\rm th}/\sqrt{3}$, the length scales order as $\RB< \RH<H\di$ (R06).  In this regime, many studies assume the atmosphere matches the disk conditions at $\RB$.  We, however, use $\RH$ as the matching radius in both this low mass and other higher mass regimes.  This choice is justified by the fact that, for hydrostatic solutions, the  density at $\RB$ exceeds the disk's background density by an order unity factor (R06).  Nevertheless, since the density change is modest, the choice of outer boundary condition has a similarly modest effect on our results.

@@

At higher planet masses where $M\pla > 3 M_{\rm th}$ and $H\di < \RH < \RB$, spherical symmetry is no longer a good approximation, due to both the vertical stratification of the disk and gap opening (see \S\ref{sec:hydro} for a discussion of neglected non-hydrostatic effects on all mass scales).

Note that while $\RH$ is the outer boundary in our structure calculations, we define planet masses to include only the mass inside the smaller of $\RB$ or $\RH$.    This conservative choice in quoting planet masses is usually a minor distinction because (when $\RB < \RH$) the gas between $\RB$ and $\RH$ is weakly compressed.

@@

Our atmosphere calculations use the standard structure equations of mass conservation, hydrostatic balance, thermal gradients, and energy conservation:
\begin{subeqnarray}
\label{eq:struct}
\frac{dm}{dr}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
\frac{dP}{dr}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
\frac{dT}{dr}&=&\nabla \frac{T}{P}\frac{dP}{dr}\slabel{eq:structc} \\
\frac{dL}{dr}&=&4 \pi r^2 \rho \left(\epsilon - \left. T {\partial S \over \partial t} \right|_m \right)\slabel{eq:structd}, 
\end{subeqnarray}
\noindent where $r$ is the radial coordinate, and $P$, $T$, $\rho$  and $L$ are the gas pressure, temperature, density and luminosity, respectively.  The enclosed mass  at radius $r$ is $m$. \Eq{eq:structc} simply defines the temperature gradient  $\nabla \equiv d \ln T/d \ln P$.  In optically thick radiative zones, radiative diffusion gives a temperature gradient
\begin{equation}
\label{eq:delrad}
\delrad \equiv \frac{3 \kappa P}{64 \pi G m \sigma T^4} L,
\end{equation}
where $\sigma$ is the Stefan-Boltzmann constant. In our models the atmosphere is optically thick throughout the whole radiative region (see \App{opacityan} for an analytic estimate of the optical depth). In convectively unstable regions, efficient convection gives an isentropic temperature gradient with $\nabla = \delad$, the adiabatic gradient

@@

In the energy equation (\ref{eq:structd}), $\epsilon$ represents all local sources of heat input, which excludes the motion of the atmosphere itself.  In stars, nuclear burning contributes to $\epsilon$.  In a protoplanetary atmosphere, dissipative drag on planetesimals contributes to $\epsilon$.  Our simplified models set $\epsilon = 0$, consistent with our neglect of planetesimal accretion luminosity at the base of the atmosphere.  The $\epsilon_{\rm g} = -T \partial S / \partial t$ term gives the energy input from gravitational contraction.\footnote{In general, any motion is accounted for by this term.  The partial time derivative is performed on shells of fixed mass.}  The partial time derivative would normally require our radial derivatives to be partial derivatives.  However, our subsequent developments will replace the local energy equation (\ref{eq:structd}) with global energy balance (see section \S\ref{cooling}), reverting the structure equations to time-independent ordinary differential equations (ODEs).

To solve the equation set (\ref{eq:struct}), an EOS is required for closure. In our study, we adopt an ideal gas law with a polytropic EOS 
\begin{subeqnarray}\label{eq:idealEOS}
P &=& \rho \Rg T \, ,\slabel{eq:idealgas} \\
P &=&K \rho^{\gamma} \, , \slabel{eq:polyEOS}%\equiv K \rho^{1/(1-\delad)}, 
\end{subeqnarray}
where $K$ is the adiabatic constant. The adiabatic index is  $\gamma = 1/(1 - \delad)$.  This work uses $\delad=2/7$ for an ideal diatomic gas (by comparison, an ideal monatomic gas has $\delad = 2/5$).  While our reference mean molecular weight ($\mu = 2.35$ proton masses) includes Helium, we ignore Helium's effect on the EOS, which is already greatly simplified. The second law or thermodynamics gives the relative entropy as
\begin{equation}
S = \Rg \ln \left(T^{1/\delad} \over P \right) ,
\end{equation} 
eliminating the need for $K$.

Boundary conditions must be satisfied at both the base and the top of the atmosphere, with $m(R\co) = M\co$, $T(\RH) = T\di$ and $P(\RH) = P\di$.  In principle, our solutions describe atmospheres with $L(R\co) = 0$. In practice, since we do not directly integrate \Eq{eq:structd} we need not directly impose this boundary condition, as described in \S\ref{sec:twolayer}.

@@

This section describes the global energy balance of a planet embedded in a gas disk, or more generally any spherical, hydrostatic object in pressure equilibrium with a background medium.  The total atmospheric energy includes gravitational and internal energies, $E = E_G + U$, with
\begin{subeqnarray}
E_G&=&-\int_{M\co}^M \frac{G m}{r} dm \, , \label{eq:Eg} \\
U&=&\int_{M\co}^M u dm \, .\slabel{eq:U}
\end{subeqnarray}
The specific internal energy is $u = C_V T = \Rg (\delad^{-1} -1) T$ for a polytropic EOS.  For a star or coreless planet, $M\co = 0$.

%AP/AY: add virial theorem here?

We start with the global energy balance for an isolated planet of mass $M$ with a free surface:
\begin{equation}
\label{eq:coolingstar}
L_M = L\co + \Gamma - \dot{E}.
\end{equation}
The surface luminosity, $L_M$, includes contributions from the core luminosity $L\co$ --  e.g.\ planetesimal accretion or radioactive decay -- from the total heat generation $\Gamma$ -- given by the integral of $\epsilon$ over the object -- and from the rate of change of atmospheric energy $\dot{E}$, a loss term.

@@

A protoplanetary atmosphere embedded in a gas disk lacks a free surface.  For objects without a free surface (or interior to a free surface), the full energy equation, 
\begin{equation}
\label{eq:coolingglobal}
L_M=L\co+\Gamma-\dot{E}+e_M \dot{M} - P_M \left. \frac{\partial V_M}{\partial t}\right|_M \, ,
\end{equation}
acquires surface terms as derived in  \App{sec:globalderiv}.  The surface can be at any mass level $M$, where the instantaneous radius is $R$.  Other surface quantities are labeled by $M$ subscripts.  The energy accreted across the surface is given by the specific energy, $e_M = u_M-G M/R$, and the mass accretion rate of gas, $\dot{M}$.  The work done by the surface is $P_M \partial V_M/ \partial t$, with the partial derivative performed at fixed mass.  %This generalized energy equation applies on any spherical shell where hydrostatic balance holds.  The choice of the mass shell $M$\footnote{\Eq{eq:coolingglobal} also applies in the interior of objects with a free surface.  The work term, $P_M \p V_M/\p t$, vanishes at the free surface.  The accretion energy, $e_M \dot{M}$,  vanishes for a truly isolated object, but would be included to account for accretion onto (or mass loss from) an otherwise free surface.}   Thus $M$ no longer refers to a uniquely defined total mass, but to the chosen mass level, where the instantaneous radius is $R$.  Values on this shell are labelled by $M$ subscripts.

@@

To simplify our calculations of atmospheric contraction, we use a two-layer model with a bottom convective region and an upper radiative layer.   The existence of such a structure is well known from previous studies (e.g., R06) and can be readily understood.  Before the protoplanetary atmosphere can cool, it has the entropy of the disk.  As the atmosphere cools, the deep interior remains convective.  Convective interiors are a common feature of low mass cool objects (brown dwarfs and planets) that results from the behavior of $\delrad$ for realistic opacity laws.  However, the entropy of the deep interior decreases as the atmosphere cools.  A region of outwardly increasing entropy, i.e.\ a radiative layer, is required to connect the convective interior to the disk.  A more complicated structure, with radiative windows in the convection zone, is possible as discussed in \S\ref{sec:op}.

@@

In the radiative zone,  local energy balance, \Eq{eq:structd}, does affect the atmospheric structure.  We proceed by assuming that the majority of energy is lost from the convective interior, and thus the luminosity can be treated as constant in the outer radiative zone.  With this approximation we can construct solutions from Equations (\ref{eq:struct}a -- c) that ``follow the mass," i.e.\ gradually increase the atmospheric mass.  We then use global energy balance, \Eq{eq:coolingglobal}, to place these solutions in a cooling sequence.  The validity of neglecting luminosity generation in the radiative zone can be checked \emph{a posteriori}.

To obtain a single atmosphere profile (indexed by $i$), we choose a planet mass $M_i$.  At the outer boundary, at $\RH(M_i)$, the temperature and pressure are set to the disk values.  To integrate Equations (\ref{eq:struct}a--c), the luminosity is required to compute $\delrad$.  The correct value of the luminosity is not known in advance, and is the eigenvalue of the problem.  The boundary conditions can only be satisfied for the correct value of the luminosity eigenvalue, which we find by the shooting method.  Specifically, we alter the luminosity until the integrated value of mass at the core, $m(R\co)$, matches the actual core mass, $M\co$.

>>>>

The core accretion hypothesis proposes that giant planets form by the accretion of gas onto a solid protoplanetary core. The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$. This value, however, depends on several factors, such as the location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids.  Motivated by ongoing direct imaging searches for giant planets, this study focuses on core mass requirements in the outer disk.
Our model investigates the growth of atmospheres around solid cores of fixed mass.  Without planetesimal accretion to heat the atmosphere, our limiting case gives the fastest possible atmospheric growth by Kelvin-Helmholtz contraction. We develop a two-layer, spherical atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk.  For different locations in the disk, we determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.   The minimum core mass required to nucleate atmosphere collapse is $\sim$$8.5 M_{\oplus}$ at 5 AU and steadily decreases to $\sim$$3.5 M_{\oplus}$ at 100 AU, for a Solar composition atmosphere with a standard dust opacity.   This trend is mainly explained by the cooler temperatures in the outer disk. Lower pressures and densities in the outer disk have a negligible effect on core mass requirements.  At all distances, decreasing the dust opacity and/or increasing the mean molecular weight of the atmosphere reduces the critical core mass.   The ongoing accretion of solids would delay or prevent atmospheric runaway growth.  Thus our low values for minimum core masses do not contradict the existence of ice giants, but emphasize the importance of the accretion history of solids.  We also develop a non-self-gravitating, analytic cooling model to help explain our numerical trends.  Additionally, this comparison reveals that self-gravity significantly affects atmospheric evolution even when the atmosphere is only $\sim$$10\%$ as massive as the core.
\end{abstract}

@@

Models of giant planet formation fall in two categories:  core accretion or gravitational instability \citep{dangelo11, youdin13}. In core accretion models, a solid core grows until it becomes massive enough to rapidly accrete gas \citep{PerCam74, mizuno78}.  Gravitational instability (GI) theories investigate the fragmentation of the protoplanetary disk into bound clumps \citep{cameron78, boss97}.

@@

This work develops quasi-static models in which the core has a fixed mass and is no longer accreting solids.  By isolating the process of KH contraction, we determine the minimum timescale for runaway atmospheric growth, and thus for giant planet formation by core accretion.  Previous studies have considered this type of growth \citep{ikoma00, pn05}, which is an end-member case of phases 2 and 3 in more detailed quasi-static models.  Especially since phase 2 atmospheric growth is often the slowest stage of core accretion, understanding the fastest allowed rates of gas accretion is crucial importance.  Our study develops simplified cooling models for the purposes of broadly exploring parameter space -- especially in the outer disk -- and developing a more detailed understanding of atmospheric accretion.

@@

Our model can  treat atmospheric growth up to the early stages of runaway growth, around the crossover mass.  At higher masses, our approximate treatment of the radiative zone (explained below) breaks down.   However, evolution after the onset of runaway growth contributes minimally to the total planet formation timescale, which we can model to good accuracy. 
%Our treatment is also inappropriate for hot, short-period planets, where dust sublimation gives deeper radiative zones that require more detailed models.

Our main assumptions are summarized as follows:
\begin{enumerate}
\item The atmosphere is spherically symmetric and remains in hydrostatic balance during its thermal evolution.
\item The core mass and radius are fixed in evolutionary calculations, neglecting ongoing planetesimal or dust accretion.
\item At the planet's Hill radius, the atmospheric temperature and pressure match the conditions in the disk midplane.
\item The only source of planetary luminosity is the gravitational contraction of the atmosphere.  
\item Cooling of the atmosphere is dominated by the convective interior.  Thus the luminosity in the radiative exterior is held spatially constant. 
\item A global cooling model connects independent static solutions into a time-dependent sequence.
\item A polytropic equation of state (EOS) is assumed for simplicity.
\item Dust grains provide the opacity in the radiative zone, which remains cool enough to avoid dust sublimation.
%\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass, i.e. the crossover time, is a good approximation of the total time to form a gas giant.
\end{enumerate}
%The remainder of this section develops our model in more detail.

@@

The characteristic length scales for protoplanetary atmospheres are crucial for choosing boundary conditions and for understanding the validity of  spherical symmetry in a gas disk of scale height $H\di$.

The solid core has a radius
\begin{equation}
\label{eq:rc}
R\co \equiv \left(\frac{3 M\co}{4 \pi \rho\co}\right)^{1/3} \approx 10^{-4} \mcn{10}^{1/3} ~\text{AU},
\end{equation}
where the core mass, $M\co$, is normalized to 10 Earth masses as $\mcn{10} \equiv M\co/(10~M_\oplus)$. The core density is held fixed at $\rho\co=3.2$ g cm$^{-3}$.  We thus neglect  the detailed equation of state of the solid core \citep{fortney07}.

A planet can bind a dense atmosphere if its escape velocity exceeds the sound speed.  This criterion is satisfied inside the Bondi radius
\begin{equation}
\label{eq:RB}
\RB \equiv \frac{G M\pla}{c\di^2} \approx 0.17 \, {\mpn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU,
\end{equation}
where the enclosed planet mass, $M\pla = M\co + M_\mathrm{atm}$, includes the core and any atmosphere within the Bondi radius and $\mpn{10} \equiv M\pla /(10~M_\oplus)$.   The contribution of the atmospheric mass is small in early evolutionary stages.

Stellar tides dominate the planet's gravity beyond the Hill radius
\begin{equation}
\label{eq:RHill}
R_{\rm H} = \left(M\pla \over 3 M_\ast \right)^{1/3}a \approx 0.22 \, {\mpn{10}^{1/3} \, \aun{10} }~\AU ,
\end{equation}
where hydrostatic balance breaks down.  (Here we use the same symbol, $M\pla$, to now mean the mass within $\RH$.)

@@

In the low mass regime, $M\pla < M_{\rm th}/\sqrt{3}$, the length scales order as $\RB< \RH<H\di$.  In this regime, many studies assume the atmosphere matches the disk conditions at $\RB$.  We use $\RH$ as the matching radius in all regimes.  We are accounting for the fact that, in hydrostatic models, the density at 
 $\RB$ is higher than the background density (see R06). Nevertheless, since the density change is modest, these two boundary conditions give similar results.

@@

At higher planet masses where $M\pla > 3 M_{\rm th}$ and $H\di < \RH < \RB$, spherical symmetry is no longer a good approximation, due to both the vertical stratification of the disk and gap opening.  See \S\ref{sec:hydro} for further discussion of neglected hydrodynamic effects.

While $\RH$ is always the outer boundary in our structure calculations, we quote planet masses as the enclosed mass within the smaller of $\RB$ and $\RH$.  %  This conservative choice in quoting planet masses is usually a minor distinction because (when $\RB < \RH$) the gas between $\RB$ and $\RH$ is weakly compressed.

@@

Our atmosphere calculations use the standard structure equations of mass conservation, hydrostatic balance, thermal gradients, and energy conservation:
\begin{subeqnarray}
\label{eq:struct}
\frac{dm}{dr}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
\frac{dP}{dr}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
\frac{dT}{dr}&=&\nabla \frac{T}{P}\frac{dP}{dr}\slabel{eq:structc} \\
\frac{dL}{dr}&=&4 \pi r^2 \rho \left(\epsilon - \left. T {\partial S \over \partial t} \right|_m \right)\slabel{eq:structd}, 
\end{subeqnarray}
\noindent where $r$ is the radial coordinate, $L$ is the luminosity, and $P$, $T$, $\rho$  and $S$ are the gas pressure, temperature, density and entropy, respectively.  The enclosed mass  at radius $r$ is $m$. \Eq{eq:structc} simply defines the temperature gradient  $\nabla \equiv d \ln T/d \ln P$.

Radiation diffusion gives a temperature gradient
\begin{equation}
\label{eq:delrad}
\delrad \equiv \frac{3 \kappa P}{64 \pi G m \sigma T^4} L\, ,
\end{equation}
with $\sigma$ the Stefan-Boltzmann constant.  In our models, optically thick diffusion is a good approximation throughout the radiative zones.  In convectively unstable regions, efficient convection gives an isentropic temperature gradient with $\nabla = \delad$, the adiabatic gradient

@@

In the energy equation (\ref{eq:structd}), $\epsilon$ represents all local sources of heat input, except for the motion of the atmosphere itself.  From stellar structure, $\epsilon$ may be familiar as a nuclear burning term.  In a protoplanetary atmosphere, dissipative drag on planetesimals contributes to $\epsilon$.  Our models set $\epsilon = 0$, consistent with our neglect of planetesimal accretion.

The energy input from gravitational contraction, $\epsilon_{\rm g} = -T \partial S / \partial t$, is crucial for a cooling model.\footnote{In general, any motion is accounted for by this term.  The partial time derivative is performed on shells of fixed mass.}  The partial time derivative would normally require our radial derivative to be partial as well.  However, our subsequent developments will replace the local energy equation (\ref{eq:structd}) with global energy balance (see section \S\ref{cooling}), reverting the structure equations to time-independent ordinary differential equations (ODEs).

To solve the equation set (\ref{eq:struct}), an EOS is required for closure. In our study, we adopt an ideal gas law with a polytropic EOS 
\begin{subeqnarray}\label{eq:idealEOS}
P &=& \rho \Rg T \, ,\slabel{eq:idealgas} \\
S &=& \Rg \ln \left(T^{1/\delad} \over P \right) \, ,\slabel{eq:polyEOS}  % P &=&K \rho^{\gamma} \, , 
\end{subeqnarray}
with $S$ a relative entropy.  This work uses $\delad=2/7$, i.e.\  a polytropic index $\gamma \equiv 1/(1 - \delad) = 7/5$, for an ideal diatomic gas.\footnote{\Eq{eq:polyEOS} is equivalent to the standard polytropic relation,  $P =K \rho^{\gamma}$, if the polytropic index $K$ replaces $S$.}    We thus neglect the presence of monatomic Helium in our choice of polytropes, a common practice in idealized studies.  We do however account for Helium in our reference value of $\mu = 2.35$ proton masses.

Boundary conditions must be satisfied at both the base and the top of the atmosphere, with $m(R\co) = M\co$, $T(\RH) = T\di$ and $P(\RH) = P\di$.  Our model atmospheres also obey $L(R\co) = 0$. Along with \Eq{eq:structd}, this boundary condition is incorporated in the global cooling model described below.

@@

We now consider the global energy balance of a planet embedded in a gas disk.  More generally our derivation applies to any spherical, hydrostatic object in pressure equilibrium with a background medium.  The total atmospheric energy includes gravitational and internal energies, $E = E_G + U$, with
\begin{subeqnarray}
E_G&=&-\int_{M\co}^M \frac{G m}{r} dm \, , \label{eq:Eg} \\
U&=&\int_{M\co}^M u dm \, .\slabel{eq:U}
\end{subeqnarray}
The specific internal energy is $u = C_V T = \Rg (\delad^{-1} -1) T$ for a polytropic EOS.  For a star or coreless planet, $M\co = 0$.

We start with a standard result, global energy balance for an isolated, i.e.\ not embedded, planet:
\begin{equation}
\label{eq:coolingstar}
L_M = L\co + \Gamma - \dot{E}.
\end{equation}
The surface luminosity, $L_M$, includes the core luminosity $L\co$ from e.g.\ planetesimal accretion or radioactive decay.   The total heat generation $\Gamma$  is the integral of $\epsilon$ over the atmosphere.  The rate of change of atmospheric energy, $\dot{E}$, is a loss term.

@@

A protoplanetary atmosphere embedded in a gas disk lacks a free surface.  For objects without a free surface (or interior to a free surface), the full energy equation, 
\begin{equation}
\label{eq:coolingglobal}
L_M=L\co+\Gamma-\dot{E}+e_M \dot{M} - P_M \left. \frac{\partial V_M}{\partial t}\right|_M \, ,
\end{equation}
acquires surface terms as derived in  \App{sec:globalderiv}.  The surface can be at any mass level $M$, where the instantaneous radius is $R$.  Surface quantities are labeled by $M$ subscripts (except for $R$).  The energy accreted across the surface is given by the specific energy, $e_M = u_M-G M/R$, and the mass accretion rate of gas, $\dot{M}$.  The work done by the surface is $P_M \partial V_M/ \partial t$, with the partial derivative performed at fixed mass.

@@

To simplify our evolutionary calculations, we develop a two-layer atmospheric model with a convective interior and a radiative exterior.   The existence of this layered structure is well known from previous studies (e.g., R06) and can be readily understood.  Before the protoplanetary atmosphere can cool, it has the entropy of the disk.  As the atmosphere cools, the deep interior remains convective.  Convective interiors are a common feature of low mass cool objects (brown dwarfs and planets) that results from the behavior of $\delrad$ for realistic opacity laws.  However, the entropy of the deep interior decreases as the atmosphere cools.  A region of outwardly increasing entropy, i.e.\ a radiative layer, is required to connect the convective interior to the disk.  A more complicated structure, with radiative windows in the convection zone, is possible as discussed in \S\ref{sec:op}.

@@

In the radiative zone,  local energy balance, \Eq{eq:structd}, does affect the atmospheric structure.  We proceed by assuming that the majority of energy is lost from the convective interior, and thus the luminosity can be treated as constant in the outer radiative zone.

With this approximation for the radiative zone, we can construct an array of solutions to Equations (\ref{eq:struct}a -- c) in order of increasing atmospheric mass.  We then use global energy balance, \Eq{eq:coolingglobal}, to ``follow the mass" and place these solutions in a cooling sequence.  The validity of neglecting luminosity generation in the radiative zone can be checked \emph{a posteriori}, see \S\ref{sec:endoftime}.

To obtain a single atmosphere solution (indexed by $i$), we choose a planet mass $M_i$.  At the outer boundary, at $\RH(M_i)$, the temperature and pressure are set to the disk values.  To integrate Equations (\ref{eq:struct}a--c), the luminosity is required to compute $\delrad$.  The correct value of the luminosity is not known in advance, and is the eigenvalue of the problem.  The boundary conditions can only be satisfied for the correct value of the luminosity eigenvalue, which we find by the shooting method.  Specifically, we alter the luminosity until the integrated value of mass at the core, $m(R\co)$, matches the actual core mass, $M\co$.
